Icicle: write once, run once

### 2016-06-02 2016fhpc/Main.tex
Change evaluator to take modality; update some benchmarks; hide conclusion
minor changes
benchmarks

\authorinfo{
  Amos Robinson$^{\alpha \beta}$
  \and Ben Lippmeier$^{\beta \gamma}$
}{
  \vspace{5pt}
  \shortstack{
    $^\alpha$Ambiata      \\
    Grow with data          \\[2pt]
    \textsf{amos.robinson@ambiata.com}
  }
  \shortstack{
    $^\beta$UNSW Australia\\
    \\[2pt]
    ~~~~\textsf{amosr,benl@cse.unsw.edu.au}
  }
  \shortstack{
    $^\gamma$Vertigo Technology \\
    \\[2pt]
    ~~\textsf{benl@vergo.co}
  }
}

@@

\begin{abstract}
We present Icicle, a pure streaming query language which statically guarantees that multiple queries over the same input stream will be fused. We use a modal type system to ensure that fused queries can be computed in an incremental fashion, and a fold-based intermediate language to compile down to efficient C code. We present production benchmarks demonstrating significant speedup over existing queries written in R, as well as the widely used Unix tools \texttt{grep} and \texttt{wc}.

@@

\input{section/01-Introduction.tex}
\input{section/02-Source.tex}
\input{section/03-Core.tex}
\input{section/04-Benchmarks.tex}
\input{section/05-Conclusion.tex}
% \input{section/XX-Elimination.tex}
\input{section/XX-Related.tex}

@@

>>>>

\authorinfo{
  Amos Robinson$^{\alpha \beta}$
  \and Ben Lippmeier$^{\beta \gamma}$
}{
  \vspace{5pt}
  \shortstack{
    $^\alpha$Ambiata      \\
    \textsf{amos.robinson@ambiata.com}
  }
  \shortstack{
    $^\beta$UNSW Australia\\
    ~~~~\textsf{amosr,benl@cse.unsw.edu.au}
  }
  \shortstack{
    $^\gamma$Vertigo Technology \\
    ~~\textsf{benl@vergo.co}
  }
}

@@

\begin{abstract}
We present Icicle, a pure streaming query language which statically guarantees that multiple queries over the same input stream will be fused. We use a modal type system to ensure that fused queries can be computed in an incremental fashion, and a fold-based intermediate language to compile down to efficient C code. We present production benchmarks demonstrating significant speedup over existing queries written in R, and on par with the widely used Unix tools \texttt{grep} and \texttt{wc}.

@@

\input{section/01-Introduction.tex}
\input{section/02-Source.tex}
\input{section/03-Core.tex}
\input{section/04-Benchmarks.tex}
% \input{section/05-Conclusion.tex}
% \input{section/XX-Elimination.tex}
\input{section/XX-Related.tex}

@@

% \input{section/A1-ExampleDerivation.tex}

### 2016-06-02 2016fhpc/section/01-Introduction.tex
benchmarks

For example queries, suppose we have a table, @stocks@, containing daily open and close prices for a set of companies. We want to compute the number of days where the open price exceeded the close price, and vice versa. We also want the mean of the open price for days in which the open price exceeded the close price. In Icicle we write the three sub queries as follows:

\begin{code}
table stocks { open : Int, close : Int }
query 
  more  = filter open > close of count;
  less  = filter open < close of count;
  mean  = filter open > close of sum open / count;
\end{code}
In the above code, @open > close@ and @close < open@ are the filter predicates, and @count@ counts the number of times the body of each filter construct was entered. We describe this construct more fully in \REF

A joint query like this can be converted to a back-end language such as SQL, but doing so by hand is tedious and error prone. As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We instead need to lift each predicate to an expression-level conditional:
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
  FROM stocks;
\end{code}
As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

@@

\item
We present production benchmarks of Icicle compiled code which compare favourably to standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
\end{itemize}

>>>>

For example queries, suppose we have a table, @stocks@, containing daily open and close prices for a set of companies. We want to compute the number of days where the open price exceeded the close price, and vice versa. We also want the mean of the open price for days in which the open price exceeded the close price. In Icicle we write the three queries as follows:

\begin{code}
table stocks { open : Int, close : Int }
query 
  more  = filter open > close of count;
  less  = filter open < close of count;
  mean  = filter open > close of sum open / count;
\end{code}
In the above code, @open > close@ and @close < open@ are the filter predicates, and @count@ counts the number of times the body of each filter construct was entered.
% We describe this construct more fully in \REF

A joint query like this can be converted to a back-end language such as SQL, but doing so by hand is tedious and error prone. As the three queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We instead need to lift each predicate to an expression-level conditional:
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
  FROM stocks;
\end{code}
As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

@@

\item
We present production benchmarks of Icicle compiled code which outperforms an existing R feature generation system by several orders of magnitude, and are on par with standard Unix utilities that conceptually perform less work\sref{s:Benchmarks}.
\end{itemize}

### 2016-06-02 2016fhpc/section/03-Core.tex
benchmarks

  after  { map : Map Date Real
               = mapMap (/) (mapZip gS gC) }
\end{code}

It is possible to convert this as a map of pairs of the sum and count.
Keeping this as two separate maps rather than a map of pairs exposes more opportunities for common subexpression elimination when fusion occurs.
For example, another query @group key of count@ can reuse the already constructed @gC@ map.

@@

If we now wish to fuse these query plans together, it is as simple as giving the @before@, @folds@ and @after@ bindings unique names and appending each part of the plan together.
The names of the returns and the column names stay the same.
Query plans can only be fused if they act on the same table.
The interesting thing here is just how simple fusion is.
Thanks to the single-pass restriction, we know there are no dependencies between any two query plans: if there were, the source query would have required multiple passes.
The single-pass restriction means that any source query can be expressed as a single fold over the data.

An efficient imperative loop can be generated from this query plan in a similar way to previous work on flow fusion~\cite{lippmeier2013data}.

>>>>

  after  { map : Map Date Real
            = mapMap (\s c. s / c) (mapZip gS gC) }
\end{code}

It would be possible to convert this as a map of pairs of the sum and count.
Keeping this as two separate maps rather than a map of pairs exposes more opportunities for common subexpression elimination when fusion occurs.
For example, another query @group key of count@ can reuse the already constructed @gC@ map.

@@

If we wish to fuse query plans on the same table together, it is as simple as giving the @before@, @folds@ and @after@ bindings unique names and appending each part of the plan together.
The names of the returns and the column names stay the same.
It is the single-pass restriction that makes fusion so simple, as it implies there are no fusion-preventing dependencies between any two query plans: if there were, the source query would have required multiple passes.

Efficient imperative loops can be generated from these query plans in a similar way to previous work on flow fusion~\cite{lippmeier2013data}.

### 2016-06-02 2016fhpc/section/04-Benchmarks.tex
benchmarks

At Ambiata, we are currently using Icicle in production over medium-sized datasets (around three hundred gigabytes uncompressed) that fit on a single disk.
These initial results have been very promising, and we are currently implementing distribution across multiple machines to handle datasets that are tens of terabytes compressed, and do not fit on a single disk.
Incremental computation is even more important for the distributed case, as it can potentially avoid copying terabytes of data over the network.

@@

The R script for this takes fifteen hours to run and is 3,566 lines, 2,311 of which are code.
In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.

@@

\begin{tikzpicture}
\begin{axis}[
	x tick label style={/pgf/number format/1000 sep=},
	ylabel=MB/s,
	enlargelimits=0.01,
	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
	ybar interval=0.7,
  symbolic x coords={R, wc, CPU1, grep, wc -l, cat, Run 372, Run 1, end}
]
\addplot coordinates {(R,6.0) (wc,182.6) (CPU1,186.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };

@@

\caption{Throughput comparisons of Icicle against existing R code and standard Unix utilities; higher is faster.}
\label{fig:bench:other}
\end{figure}

@@

\TODO{update to talk about single CPU number.
Single CPU version 30 times faster than R, and in the same ballpark as @wc@ and @grep@.
Because the memory usage is bounded, we have a better idea of how many we can run on a single machine without running out of memory and stepping on each other's toes.
This means we can scale up to get quite close to the disk speed.
This is not true of the R code: assuming perfect scaling, we would need at least 150 threads.
Furthermore, the single-threaded R script is already using all the memory, and it is not obvious whether memory usage will decrease linearly as the input is divided and more threads are added.
Blah~\cite{mcsherry2015scalability}
}

The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
We compared the throughput of several programs over the same 317GB dataset:
\begin{itemize}
\item our original R implementation (R);
\item Icicle with equivalent queries (Run 372);
\item Icicle with a single query computing the mean (Run 1);
\item finding empty lines with @grep "^$"@;
\item counting characters, words and lines with @wc@;
\item counting only lines with @wc -l@; and
\item reading and throwing away the results with @cat > /dev/null@.
\end{itemize}

We ran all of the unix utilities with unicode disabled using @LANG=C LC_COLLATE@ for maximum performance.
The computer we used for most of these was an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
The R code requires an @i2.8xlarge@ with 244GB of memory, but we were unable to perform our other benchmarks on such a large machine.
Icicle significantly outperformed R, @grep@ and @wc@, and was roughly on par with @wc -l@ and @cat@.
Although the computer had 32 CPUs, Icicle was the only benchmark we ran that used multiple CPUs: the other benchmarks were single threaded.

Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.
We ran with a number of queries twice, with one version writing output to disk, and the other throwing away the result to @/dev/null@ rather than spending time writing.
The graph shows the disk version decreasing roughly linearly in the number of queries, while ignoring the output stays constant.
This suggests that the output code is the bottleneck, which is unsurprising given the output format is text-based PSV.
The upshot of this is that the time spent computing the queries appears constant as hundreds of queries are added, which suggests that many more queries can be handled.

>>>>

At Ambiata, we are currently using Icicle in production over medium-sized datasets that fit on a single disk.
These initial results have been very promising, and we are currently implementing distribution across multiple machines to handle datasets that are tens of terabytes compressed, and do not fit on a single disk.
Incremental computation is even more important for this distributed case, as we can avoid copying terabytes of data over the network.

@@

The R script for this takes fifteen hours to run and is 3,566 lines of code.
In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.

@@

\begin{tikzpicture}
\begin{axis}[
	x tick label style={/pgf/number format/1000 sep=},
	ylabel=MB/s,
	enlargelimits=0.01,
	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
	ybar interval=0.7,
  symbolic x coords={R, wc, 1 CPU, grep, cat, 32 CPU, end}
]
\addplot coordinates {(R,6.0) (wc,182.6) (1 CPU,186.6) (grep,261) (cat,878) (32 CPU,908) (end,0) };
% \addplot coordinates {(R,6.0) (wc,182.6) (CPU1,186.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };

@@

\caption{Throughput comparisons of Icicle (1 CPU and 32 CPU) against existing R code and standard Unix utilities; higher is faster.}
\label{fig:bench:other}
\end{figure}

@@

The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
We compared the throughput of several programs over the same 317GB dataset:
\begin{itemize}
\item our original R implementation (R);
\item Icicle running single-threaded (1 CPU);
\item Icicle running on multiple processors (32 CPU);
\item finding empty lines with @grep "^$"@;
\item counting characters, words and lines with @wc@;
\item reading and throwing away the results with @cat > /dev/null@.
\end{itemize}

We ran all of the Unix utilities with unicode disabled using @LANG=C LC_COLLATE@ for maximum performance.
The computer we used for most of these was an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
The R code requires an @i2.8xlarge@ with 244GB of memory, but we were unable to perform our other benchmarks on such a large machine.
Icicle significantly outperformed R, and the single-threaded version was on par with @wc@, while only a little slower than @grep@.
This is despite doing conceptually more work than @wc@ and @grep@.
By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
These results give us confidence that our distributed implementation will be performant as well as scalable~\cite{mcsherry2015scalability}.

Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.
We ran two versions with each number of queries; one version writing output to disk, and the other throwing away the result to @/dev/null@ rather than spending time writing.
The graph shows the disk version decreasing roughly linearly in the number of queries, while the version ignoring the output stays constant.
This suggests that the output code is the bottleneck, which is unsurprising given the output format is text-based PSV.
The upshot of this is that the time spent computing the queries appears constant as hundreds of queries are added, which suggests that many more queries can be handled.
